---
title: Rainfall
author: nickforr
date: '2017-07-09'
slug: rainfall
categories: 
  - R
tags: 
  - ggplot2
  - tidyverse
description: ''
draft: true
topics:
  - R
---

## Wetter in the West (of Scotland)?

I grew up in [Fife](https://en.wikipedia.org/wiki/Fife) and now live on the
outskirts of Glasgow.  And maybe it's just the 'lovely' July weather we're
currently experiencing, or maybe it's just that I'm getting older and have a
selective memory for the past, but I don't recall summers being quite so wet
when I was growing up!

I thought I'd take the opportunity to explore the data and play around with some
visualisation...

### Data from Met Office

The Met Office provides [historic
data](http://www.metoffice.gov.uk/public/weather/climate-historic/#?tab=climateHistoric)
from weather stations across the UK and the stations at Leuchars (in Fife) and
Paisley (just to the west of Glasgow) have data going back to the 1950s.

```{r setup}
library(tidyverse)
library(stringr)
library(viridis)
library(hrbrthemes)
```

```{r fetchData}
#Data is a bit messy in txt format (although they do have a help page on how to use the Excel 'text to columns wizard' to import the data...!)
leucharsRawData <- 
  readr::read_lines("http://www.metoffice.gov.uk/pub/data/weather/uk/climate/stationdata/leucharsdata.txt", skip = 5) %>% #skip first five lines of explanations
  stringr::str_trim() %>% #remove whitespace
  stringr::str_split(stringr::boundary("word")) %>% #split lines into columns
  purrr::discard(~("Provisional" %in% .x)) #Remove recent provisional results
paisleyRawData <- 
  readr::read_lines("http://www.metoffice.gov.uk/pub/data/weather/uk/climate/stationdata/paisleydata.txt", skip = 5) %>%
  stringr::str_trim() %>% #remove whitespace
  stringr::str_split(stringr::boundary("word")) %>% #split lines into columns
  purrr::discard(~("Provisional" %in% .x)) #Remove recent provisional results

#Check that column names are the same
stopifnot(leucharsRawData[[1]] == paisleyRawData[[1]])

#Note that 2nd row of each resulting list is an unhelpful second row of column headers
leucharsData <- 
  tail(leucharsRawData, -2) %>%
   purrr::map_df(~set_names(as.list(as.numeric(.x)), leucharsRawData[[1]])) %>%
  dplyr::mutate(dataDate = lubridate::dmy(paste(01, mm, yyyy, sep = "_"))) %>%
  select(-yyyy, -mm)
paisleyData <- 
  tail(paisleyRawData, -2) %>%
   purrr::map_df(~set_names(as.list(as.numeric(.x)), paisleyRawData[[1]])) %>%
  dplyr::mutate(dataDate = lubridate::dmy(paste(01, mm, yyyy, sep = "_"))) %>%
  select(-yyyy, -mm)

allData <- 
  list(Leuchars = leucharsData, Paisley = paisleyData) %>%
  dplyr::bind_rows(.id = "station") %>%
  dplyr::filter(
    dataDate >= max(min(leucharsData$dataDate), min(paisleyData$dataDate)))
```








